---
title: "STA302 Final project"
output:
  pdf_document: default
  html_document: default
date: "2024-11-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

install.packages("car")
library(car)
# Read the CSV file into a data frame
data <- read.csv("nba_final.csv")
new <- subset(as.data.frame(data), select = c(PTS, MP, GS, G, FG., AST, Play))


# Summaries for key variables
summary(new$PTS)  # Points per Game
summary(new$FG.)  # Field Goal Percentage
summary(new$AST)  # Assists
summary(new$G)    # Games Played
summary(new$GS)   # Games Started
summary(new$MP)   # Minutes Played

# Categorical variable Play summary
play_summary <- table(data$Play)
play_summary

```

## Fitting the model

```{r cars}
# Fit the initial model
model <- lm(PTS ~ MP + GS + G + FG. + AST + Play, data = new)
summary(model)

# Residuals vs Fitted plot
y_value <- resid(model)
x_value <- fitted(model)
plot(x = x_value, y = y_value, main="Residual vs Fitted", xlab="Fitted", ylab="Residuals")

```

## Applying a square root transformation to the response variable

```{r pressure, echo=FALSE}
# Load necessary packages
#install.packages("car")
#library(car)

# Fit a model with square root transformed PTS
sqrtPPG <- sqrt(new$PTS)
model2 <- lm(sqrtPPG ~ MP + GS + G + FG. + AST + Play, data = new)

# Residual vs Fitted plot for transformed model
x_value1 <- fitted(model2)
y_value1 <- resid(model2)
par(mfrow=c(1,3))
plot(x = x_value1, y = y_value1, main="Residual vs Fitted", xlab="Fitted", ylab="Residuals")

# Q-Q plots
qqnorm(y_value, main="Original (y)")
qqline(y_value)

qqnorm(y_value1, main="Square Root Transformed (sqrt(y))")
qqline(y_value1)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


## Residuals vs predictors

```{r pressure, echo=FALSE}

new <- new[complete.cases(new), ]

# Define a custom layout: 2 rows, 5 columns
layout(matrix(1:10, nrow = 2, ncol = 5, byrow = TRUE))

# Residuals vs Quantitative Predictors (New Model)
plot(new$MP, y_value1, main = "New: Residuals vs MP", xlab = "Minutes Played (MP)", ylab = "Residuals")
plot(new$G, y_value1, main = "New: Residuals vs G", xlab = "Games Played (G)", ylab = "Residuals")
plot(new$AST, y_value1, main = "New: Residuals vs AST", xlab = "Assists (AST)", ylab = "Residuals")
plot(new$GS, y_value1, main = "New: Residuals vs GS", xlab = "Games Started (GS)", ylab = "Residuals")
plot(new$`FG.`, y_value1, main = "New: Residuals vs FG%", xlab = "Field Goal Percentage (FG%)", ylab = "Residuals")

# Residuals vs Quantitative Predictors (Old Model)
plot(new$MP, y_value, main = "Old: Residuals vs MP", xlab = "Minutes Played (MP)", ylab = "Residuals")
plot(new$G, y_value, main = "Old: Residuals vs G", xlab = "Games Played (G)", ylab = "Residuals")
plot(new$AST, y_value, main = "Old: Residuals vs AST", xlab = "Assists (AST)", ylab = "Residuals")
plot(new$GS, y_value, main = "Old: Residuals vs GS", xlab = "Games Started (GS)", ylab = "Residuals")
plot(new$`FG.`, y_value, main = "Old: Residuals vs FG%", xlab = "Field Goal Percentage (FG%)", ylab = "Residuals")

# Power transform on predictors

# Adjust predictors to ensure positivity if necessary
new$GS_adj <- new$GS + 1  # Games Started
new$FG_adj <- new$`FG.` + 1  # Field Goal Percentage
new$AST_adj <- new$AST + 1  # Assists
new$MP_adj <- new$MP + 1  # Minutes Played
new$G_adj <- new$G + 1  # Games Played

# Run power transformation on the predictors
power_results <- powerTransform(cbind(new$G_adj, new$GS_adj, new$MP_adj, new$FG_adj, new$AST_adj))

# New model with transformed predictors
# Apply power transformations to each predictor based on the Estimated Power
new$G_transformed <- new$G_adj^1.2345
new$GS_transformed <- new$GS_adj^0.1283
new$MP_transformed <- new$MP_adj^0.8316
new$FG_transformed <- new$`FG_adj`^1.2727
new$AST_transformed <- new$AST_adj^(-0.5961)

new$sqrtPPG <- sqrt(new$PTS)

model3 <- lm(sqrtPPG ~ G_transformed + GS_transformed + MP_transformed + FG_transformed + AST_transformed + Play, data = new)

# Extract residuals from the model3
residuals_model3 <- resid(model3)

# Set up a 1-row, 5-column layout for the plots
par(mfrow = c(1, 5))

# Residuals vs G (Games Played)
plot(new$G_transformed, residuals_model3, 
     main = "Residuals vs G (Transformed)", 
     xlab = "Games Played (Transformed)", 
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

# Residuals vs GS (Games Started)
plot(new$GS_transformed, residuals_model3, 
     main = "Residuals vs GS (Transformed)", 
     xlab = "Games Started (Transformed)", 
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

# Residuals vs MP (Minutes Played)
plot(new$MP_transformed, residuals_model3, 
     main = "Residuals vs MP (Transformed)", 
     xlab = "Minutes Played (Transformed)", 
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

# Residuals vs FG% (Field Goal Percentage)
plot(new$FG_transformed, residuals_model3, 
     main = "Residuals vs FG% (Transformed)", 
     xlab = "Field Goal Percentage (Transformed)", 
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

# Residuals vs AST (Assists)
plot(new$AST_transformed, residuals_model3, 
     main = "Residuals vs AST (Transformed)", 
     xlab = "Assists (Transformed)", 
     ylab = "Residuals")
abline(h = 0, col = "red", lty = 2)

# Boxplot for categorical variable Play
boxplot(y_value ~ new$Play , main="Residual vs All Star Appearance",
xlab="All Star Appearance", ylab="Residuals")


```

## Updated model

```{r pressure, echo=FALSE}
# Model 3: transformed predictors

model3 <- lm(sqrtPPG ~ G_transformed + GS_transformed + MP_transformed + FG. + AST + Play, data = new)

y_value3 <-resid(model3)
x_value3 <-fitted(model3)
plot(x = x_value3, y = y_value3, main="Residual vs Fitted ", xlab="Fitted",ylab="Residuals") 

qqnorm(y_value3, main="sqrt(y) and Transforjme d\\\ Predictors")
qqline(y_value3)

```

## Checking additional conditions
```{R}
# Condition 1: Conditional mean response
model4 <- lm(sqrtPPG ~ GS_transformed + MP_transformed + FG. + AST + Play, data = new)
x_value4 <- fitted(model4)
y_value4 <- resid(model4)


plot(x = x_value4, y_value4 = sqrtPPG, main="Points per game vs Fitted", xlab="Fitted", ylab="Points per game")
abline(a = 0, b = 1, lty = 2)

# Condition 2: Conditional mean predictors
pairs(new[, c(2,3,5,6)])
cor(new[, c(2,3,5,6)])
```
## Hypothesis testing
```{R}
summary(model4)

#Confidence interval of the transformed model
confint(model4)

# The F-statistic of 2167 with a p-value < 2.2e-16 in your model summary indicates that the overall regression model is highly significant.
```

## Partial F-test
```{R}

anova(model4, model3)

# We decide to keep G_transformed and keep model 3
```
## Assesing multicollinearity: VIF 
```{R}
vif(model3)
vif(model4)
# We use model 3 because anova says G is significant and vif is only marginally above 5
```

#Identifying Problematic Observations
```{R}
dfbetas <- dfbetas(model3)
dim(dfbetas)
#given dimensions: 1404, 6

n=1404

cutoff_dfbetas <- 6/sqrt(n)
which(abs(dfbetas[,1])>cutoff_dfbetas)
which(abs(dfbetas[,2])>cutoff_dfbetas)

hii <- hatvalues(model3)
cutoff_hii <- 2*6/n
which(hii > cutoff_hii)

ri <- rstandard(model3)
which(ri > 2 | ri < -2)

di <- cooks.distance(model3)
cutoff_di <-qf(0.5, 2, 35 -2)
which(di>cutoff_di)

dffits <- dffits(model3)
cutoff_dffits <- 2*sqrt(6/n)
which(abs(dffits) > cutoff_dffits)

# Identify flagged observations for each diagnostic test
flagged_dfbetas_1 <- which(abs(dfbetas[, 1]) > cutoff_dfbetas)  # For Intercept
flagged_dfbetas_2 <- which(abs(dfbetas[, 2]) > cutoff_dfbetas)  # For another predictor
flagged_hii <- which(hii > cutoff_hii)                         # High leverage
flagged_ri <- which(ri > 2 | ri < -2)                          # Standardized residuals
flagged_di <- which(di > cutoff_di)                            # Cook's Distance
flagged_dffits <- which(abs(dffits) > cutoff_dffits)           # DFFITS

# Combine all flagged observations into a single vector
all_flagged <- c(flagged_dfbetas_1, flagged_dfbetas_2, flagged_hii, flagged_ri, flagged_di, flagged_dffits)

# Count the frequency of each observation being flagged
flagged_counts <- table(all_flagged)

# Find observations flagged by two or more diagnostics
multi_flagged <- as.numeric(names(flagged_counts[flagged_counts >= 2]))

# Print the observations flagged by two or more tests
print(multi_flagged)

# Optionally, inspect the corresponding rows in the dataset
problematic_data <- new[multi_flagged, ]
print(problematic_data)

## Explain why PTS (left skewed) explain why outliers should be kept.
```

```{R}
# Cook's Distance
plot(di, main = "Cook's Distance", xlab = "Observation Index", ylab = "Cook's Distance")
abline(h = cutoff_di, col = "red", lty = 2)

# Hat Values
plot(hii, main = "Hat Values", xlab = "Observation Index", ylab = "Hat Values")
abline(h = cutoff_hii, col = "red", lty = 2)

# DFFITS
plot(dffits, main = "DFFITS", xlab = "Observation Index", ylab = "DFFITS")
abline(h = c(-cutoff_dffits, cutoff_dffits), col = "red", lty = 2)

```

```{R}
#Step AIC
library(MASS)
stepAIC(lm(sqrtPPG ~ G_transformed + GS_transformed + MP_transformed + FG. + AST + Play, data=new[,-1]),
scope=list(lower=lm(sqrtPPG ~ 1, data=new[,-1])),
direction = "backward", k=2)

#Step BIC
stepAIC(lm(sqrtPPG ~ 1, data=new[,-1]),
scope=list(upper=lm(sqrtPPG ~ G_transformed + GS_transformed + MP_transformed + FG. + AST + Play, data=new[,-1])),
direction = "forward", k=log(nrow(new)))

#All Possible Subsets
install.packages("leaps")
library(leaps)
library(car)
# runs all possible subsets
best <- regsubsets(sqrtPPG ~ G_transformed + GS_transformed + MP_transformed + FG. + AST + Play, data=new[,-1], nbest=1, nvmax=7)
summary(best)
```
```



